\bf Estimating the number of clusters using cross-validation

### 2019-06-30 JCGS-R3/appendix.tex
fix appendix bib refs
fix remaining refs

It will suffice to prove self-consistency for a single fold of the
cross-validation procedure.  As in section~\ref{sec:gabriel-cv-algorithm} we
assume that the $P$ variables of the data set have been partitioned into $p$
predictor variables represented in vector~$X$ and $q$ response variables
represented in vector~$Y$.  The $N$ observations have been divided into two
sets: $n$ train observations and $m$ test observations.  We state the
assumptions for the self-consistency result in terms of a specific split; for
the result to hold in general, with high probability, these assumptions would
have to hold with high probability for a random split. The following theorem
gives conditions for Gabriel cross-validation to recover the true number of
clusters in the absence of noise.

@@

Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $\CV(k)$ be the cross-validation error for
this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
We will assume that there are $K$ ``true cluster centers''
$\mu(1), \dotsc,\mu(K)$, with
the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
\begin{enumerate}[label=(\roman*)]
  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.
  \item \label{asn:self-consistency-distinct-mux}
    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
  \item \label{asn:self-consistency-distinct-muy}
    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
  \item \label{asn:self-consistency-train}
    The training set contains at least one member of each cluster: for all $g$
    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
    $1, \dotsc, n$ such that $G_i = g$.
  \item \label{asn:self-consistency-test}
    The test set contains at least one member of each cluster: for all $g$ in
    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
    \dotsc, n+m$ such that $G_i = g$.
\end{enumerate}
Then $\CV(k) > \CV(K)$ for $k < K$, and $\CV(k) = \CV(K)$ for $k > K$, so that
Gabriel cross-validation correctly chooses $k = K$.
\end{proposition}

@@

One might object to the condition~\eqref{eqn:clust-anyd} being stated in terms
of a particular $(X, Y)$ split of the data.  In practice, one might want the
condition to hold for all possible splits of the data. A sufficient guarantee
that~\eqref{eqn:clust-anyd} holds for any $(X, Y)$ split of the data is that
the data covariance matrix $\Sigma$ is diagonal. The ad-hoc adjustment we
propose in Sec.~\ref{sec:corr-correct} attempts to transform the data to this
form.

@@

Similar as in the proof of Proposition~\ref{prop:single-general-dim}, by the variance and bias decomposition of MSE, the variance is the same, so only the bias influences the result. Note the predicted center is grand $0$ for $CV(1)$, so to see if  $CV(2) < CV(1)$ one only need to see if $||\bar{\mu}^Y_1-\hat{\mu}^Y_1||^2 < ||\hat{\mu}^Y_1 - 0||^2$, which is true if 
\[ 2\Phi(\lambda_y) + 2 \frac{\varphi(\lambda_y)}{\lambda_y} < 4\Phi(\lambda_x)-1 \] 
this result reduces to the original result of Proposition~\ref{prop:twoclust}
if one sets $\lambda_x = \mu^X$ and $\lambda_y = \mu^Y$.
\end{proof}

@@

Figure~\ref{fig:5_clusters} provides a more in-depth comparison with the
Tavazoie clusters, using a graphical confusion matrix.  The plot in cell
$(i,j)$ of the upper left part of this figure gives the mean expression level
for genes in the intersection of Tavazoie's Cluster~$i$ and our Cluster~$j$;
the plots in the margins give the mean expression levels for Tavazoie's
clusters (top right) and our clusters (bottom left). In
Figure~\ref{fig:5_clusters}, we only include a plot for cell $(i,j)$ if the
number of genes in that cell is greater than $20$.

@@

>>>>

It will suffice to prove self-consistency for a single fold of the
cross-validation procedure.  As in section~2.2 of the main article, we
%\ref{sec:gabriel-cv-algorithm} 
assume that the $P$ variables of the data set have been partitioned into $p$
predictor variables represented in vector~$X$ and $q$ response variables
represented in vector~$Y$.  The $N$ observations have been divided into two
sets: $n$ train observations and $m$ test observations.  We state the
assumptions for the self-consistency result in terms of a specific split; for
the result to hold in general, with high probability, these assumptions would
have to hold with high probability for a random split. The following theorem
gives conditions for Gabriel cross-validation to recover the true number of
clusters in the absence of noise.

@@

Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $\CV(k)$ be the cross-validation error for
this fold, computed as described in Section~2.2 of the main article.
%\ref{sec:gabriel-cv-algorithm}.
We will assume that there are $K$ ``true cluster centers''
$\mu(1), \dotsc,\mu(K)$, with
the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
\begin{enumerate}[label=(\roman*)]
  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.
  \item \label{asn:self-consistency-distinct-mux}
    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
  \item \label{asn:self-consistency-distinct-muy}
    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
  \item \label{asn:self-consistency-train}
    The training set contains at least one member of each cluster: for all $g$
    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
    $1, \dotsc, n$ such that $G_i = g$.
  \item \label{asn:self-consistency-test}
    The test set contains at least one member of each cluster: for all $g$ in
    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
    \dotsc, n+m$ such that $G_i = g$.
\end{enumerate}
Then $\CV(k) > \CV(K)$ for $k < K$, and $\CV(k) = \CV(K)$ for $k > K$, so that
Gabriel cross-validation correctly chooses $k = K$.
\end{proposition}

@@

One might object to the condition~\eqref{eqn:clust-anyd} being stated in terms
of a particular $(X, Y)$ split of the data.  In practice, one might want the
condition to hold for all possible splits of the data. A sufficient guarantee
that~\eqref{eqn:clust-anyd} holds for any $(X, Y)$ split of the data is that
the data covariance matrix $\Sigma$ is diagonal. The ad-hoc adjustment we
propose in Section~4 of the main article%.~\ref{sec:corr-correct} 
attempts to transform the data to this form.

@@

Similar as in the proof of Proposition~\ref{prop:single-general-dim}, by the variance and bias decomposition of MSE, the variance is the same, so only the bias influences the result. Note the predicted center is grand $0$ for $CV(1)$, so to see if  $CV(2) < CV(1)$ one only need to see if $||\bar{\mu}^Y_1-\hat{\mu}^Y_1||^2 < ||\hat{\mu}^Y_1 - 0||^2$, which is true if 
\[ 2\Phi(\lambda_y) + 2 \frac{\varphi(\lambda_y)}{\lambda_y} < 4\Phi(\lambda_x)-1 \] 
this result reduces to the original result of Proposition~2
of the main article %\ref{prop:twoclust}
if one sets $\lambda_x = \mu^X$ and $\lambda_y = \mu^Y$.
\end{proof}

@@

Figure~8 of the main article %\ref{fig:5_clusters}
provides a more in-depth comparison with the
Tavazoie clusters, using a graphical confusion matrix.  The plot in cell
$(i,j)$ of the upper left part of this figure gives the mean expression level
for genes in the intersection of Tavazoie's Cluster~$i$ and our Cluster~$j$;
the plots in the margins give the mean expression levels for Tavazoie's
clusters (top right) and our clusters (bottom left). In
Figure~8 (main article), %\ref{fig:5_clusters},
we only include a plot for cell $(i,j)$ if the
number of genes in that cell is greater than $20$.

@@

\bibliographystyle{apalike}
\bibliography{references}

### 2019-06-30 JCGS-R3/cvclust.tex
fix remaining refs

For the clusters that Tavazoie et al.\ were able to characterize, our analysis
broadly agrees with the earlier clustering (Section~8 of the supplementary
appendix contains a more detailed comparison). The major difference between our
analysis and that of \citet{tavazoie1999systematic} is that we are able to
identify meaningful groups of genes with a much smaller value of $k$ ($k = 5$
instead of $k = 30$), and we are able to interpret all of the clusters
found by our analysis.

>>>>

For the clusters that Tavazoie et al.\ were able to characterize, our analysis
broadly agrees with the earlier clustering. We can see this in
Figure~\ref{fig:5_clusters}, which gives us a graphical confusion matrix
comparing the two analyses.  The plot in cell $(i,j)$ of the upper left part
of this figure gives the mean expression level for genes in the intersection
of Tavazoie's Cluster~$i$ and our Cluster~$j$; the plots in the margins give
the mean expression levels for Tavazoie's clusters (top right) and our
clusters (bottom left).  We only include a plot for cell $(i,j)$ if the number
of genes in that cell is greater than $20$.  Section~8 of the supplementary
appendix contains a more detailed comparison.
The major difference between our analysis and that of
\citet{tavazoie1999systematic} is that we are able to identify meaningful
groups of genes with a much smaller value of $k$ ($k = 5$ instead of $k =
30$), and we are able to interpret all of the clusters found by our analysis.
